Journey: A Study of Studying Large Memory Systems

### 2017-04-03 paper.tex

@@

>>>>

\section{Introduction}

Recent advances in memory technologies are making it possible for machines with
large amounts of memory to become commonplace. Many current systems already
have dozens or hundreds of gigabytes of memory. These technology advances can
enable systems with terabytes of memory in the not very distant future. For
example, non-volatile memory technologies, which have denser physical
structures than traditional DRAM are making such large memories available and
more accessible \cite{xpoint}.

Much research is currently being done to study how non-volatile and large
memory systems would change system design. HP has been working on a system
which it calls ``The Machine'', in which hundreds of cores have access to a
multi-petabyte shared memory pool \cite{hp_machine}. Intel and
Micron have announced their 3D Xpoint non-volatile memory technology, making
dense, fast, persistent memory comercially available \cite{xpoint}.

However, studies of large-memory systems are often forced to use simulation or
emulation to evaluate solutions because most researchers do not have access to
these sorts of systems for experimentation \cite{quartz}. This often
forces researchers to simplify models, incurring inaccuracy, or to use smaller
benchmarks, which are less realistic. We note that there was a similar problem
when large storage systems first became available \cite{david, exalt}.

Currently, there are few good tools available to systems researchers for
understanding how systems interact with large memories. The ideal end-goal of
this research is to identify bottlenecks in managing large memories and find a
general lightweight technique or tool for studying and designing operating
systems capable of handling large memory while using systems currently
available to researchers.

@@

\section{Methodology}

Our methodology is broken into two parts. First, we measure the scalability of
several Linux memory management system calls on a system running natively on
bare metal. Then, we measure the scalability of Linux virtual machines running
on the KVM hypervisor.

\subsection{Native Experiments}

\begin{figure}
\centering
\begin{tabular}{|l|p{5cm}|} \hline
OS & Ubuntu 16.04.2 \\ \hline
Kernel & Linux 4.4.0-70 \\ \hline
CPU & Intel Xeon E5645, 2.40 GHz, 6 cores/12 threads, 384KB Private L1-I Cache,
1.536 Unified L2 Cache, 12MB Shared L3 Cache \\ \hline
Memory & 16GB, 1066MHz \\ \hline
Storage & 500GB, 7200RPM \\
\hline
\end{tabular}
\caption{Specifications of our test machine.  \label{fig:specs}}
\end{figure}

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space. (TODO: what about page tables? struct page?)
When a program makes a memory management system calls such as \texttt{mmap},
\texttt{mremap}, \texttt{munmap}, \texttt{mprotect}, \texttt{mlock}, or
\texttt{madvise}, the kernel is creating, changing, or removing
\texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux memory
management requires measuring the overhead of these structs.

We run several different microbenchmarks, each of which has a different memory
allocation pattern. These benchmarks are designed to stress the memory
management system in different ways, rather than simulate real workloads. For
each workload, we measure the latency and increase in memory usage due to each
operation in the benchmark. Notably, we do not touch the pages that we allocate,
avoiding both a page fault and the allocation of a back physical page for the
allocated virtual pages. This allows us to measure just the overhead of memory
management, as opposed to physical page allocation, swapping, page faults, or
other overheads.  Each workload contains $2^{20}$ memory management operations.

The specifications of the machine are listed in Figure \ref{fig:specs}. We
disabled Intel SpeedStep and TurboBoost. All experiments are pinned to the same
core on our machine. Each benchmark was run 5 times and the results aggregated
as described below. We reboot the machine when switching to a different
benchmark, but keep the machine running continually during the runs of the same
benchmark. Each run is executed in isolation, without any additional processes
running beyond services that start when the system is booted and a single ssh
session. We are careful to avoid opening extra ssh sessions, running screen or
tmux, or executing extra commands while a benchmark is running, as these may
impact memory usage, and thus our results.

We measured latency using the \texttt{rdtsc} instruction provided by x86\_64
processors, which gives high resolution cycle-level timestamps. We take a
timestamp before and after each memory management operation in the workload and
take the difference to get latency. The median latency of each operation is
reported across all 5 runs.

To measure memory, we use the procfs's reporting on current memory usage, along
with the slab allocator's count of active \texttt{vm\_area\_struct}s in the
system. Measuring kernel memory usage and associating memory usage with
processes is very difficult. Our methodology assumes that our benchmarks running
in otherwise idle systems should dominate increases in memory usage.  Thus, we
compute total memory usage in the system as the total amount of memory less the
amount of memory used. From this amount, we subtract the resident set size of
the benchmark process. The result is the memory usage of the rest of the system
$R$. We denote $R_i$ as the $i$-th measurement of $R$ during a run ($i = 0, 1,
... 2^{20}$). System memory usage is known to jitter in Linux (TODO: cite). To
adjust for this jitter, we define $\Delta_i = R_i - R_0$. For each operation, we
take the median value of $\Delta_i$ across the 5 runs to reduce jitter further.
This is the value graphed in the figures in this section. $\Delta_i$ represents
the increase in memory usage due to memory management overheads caused by
running $i$ memory management operations.

The workloads we measure are

\begin{itemize} \item \textit{Control}. This workload is only run for the memory
overhead experiments. It simply prints the $\Delta_i$ for each $i$. It's graph
is a straight line with positive slope. This memory usage results from the
buffer containing stdout, where our readings are being printed. All of our
experiments have the same stdout overhead, so this benchmark serves a baseline
for comparison.

\item \textit{Continuous} (``cont'' for short). This workload allocates $2^{20}$
adjacent 16KB blocks.

\item \textit{Strided}. This workload allocates $2^{20}$ untouching 16KB regions
in order of increasing memory address.

\item \textit{Random}. This workload allocates $2^{20}$ randomly placed,
non-overlaping memory regions.

\item \textit{Fragmented} (``frag'' for short). This workload first allocates
$2^{19} + 2^{18}$ contiguous 4KB blocks. Then, it resizes every other block to
40KB. This represents the worst case behavior we could think of. It requires the
kernel to find a block it has not touched in over $2^{19}$ operations and move
it somewhere else.  \end{itemize}

\subsection{VM Experiments}

TODO
